Light and life form interact to shape stomatal ratio among British angiosperms$^1$

### 2017-07-13 ms/ms.Rnw
edited ms

% Things to do before submission:
% Sally's notes
% Matt's notes
% Remove code to copy refs
% Github DOI?
% Finish abstract
% Finish citations (add citations only mentioned online to MS)

% Notes from call with Rob
% van Tienderen - sensitivity approach better for evolution
% send list of species to rob

% TASKS
% what's going on with monocots?
% in methods - mention stuff about Salisbury
% leaf size

@@

% NOTES , for example, two herbaceous species adapted to different light conditions may not really be comparable since light has a strong impact on life history selection \citep{Franco_Silvertown_1996, Galloway_Etterson_2007}.

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}). Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals did not overlap zero (Fig.~\ref{fig:SRmultReg}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distibuted across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

@@

Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio. The contributions of adaxial density, abaxial density, and their covariance are \Sexpr{round(varSDad / varSR, 2)}, \Sexpr{round(varSDab / varSR, 2)}, and \Sexpr{round(-2 * covSD / varSR, 2)}, respectively. Recall that values can be greater than one for adaxial stomatal density and negative for the covariance when the latter value is positive. This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.

Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}). Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ \Sexpr{pval2latex(exp(pvalSEM["logSDad ~ ellenberg_light"]))} and \Sexpr{pval2latex(exp(pvalSEM["logSDab ~ ellenberg_light"]))}, respectively). However, the regression of \el~on $\mathrm{sd_{ad}}$ was \Sexpr{round(fitSEM@ParTable$est[2] / fitSEM@ParTable$est[1], 1)}$\times$ that of \el~on $\mathrm{sd_{ab}}$ (\Sexpr{round(fitSEM@ParTable$est[2], 2)} versus \Sexpr{round(fitSEM@ParTable$est[1], 2)}). Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a \Sexpr{round(exp(fitSEM@ParTable$est[2]), 2)}-fold change in adaxial stomatal density versus a \Sexpr{round(exp(fitSEM@ParTable$est[1]), 2)}-fold change in abaxial stomatal density. The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $\Sexpr{pval2latex(exp(pvalSEM["logSDab ~~ logSDad"]))}). These results together imply that total stomatal density increases with \el, but the response is mediated mostly by adaxial stomatal density.

@@

The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance is not clear. Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work. Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}. However, none of these studies have accounted for the fact thats light and growth form are often confounded -- open, high light habitats are necessarily dominated by herbs -- or the fact that species are not independent because of shared evolutionaey hisory. Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods. As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Raunki\ae r life form; Fig.~\ref{fig:lf-light}). Nevertheless, both \el~and Raunki\ae r life form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms. These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.

@@

Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy. In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}). Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesies \textit{and} photosynthesis strongly limits fitness. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g. containing both annual and perennial forms). I predict that amphistomy will be favored much more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991}.

The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}. It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of pallisade mesophyll \citep{Buckley_etal_2015}. Since evaporative demand increases under high insolation, under these hypotheses we would expect plants in high light to be hypostomatous. Because stomatal conductances on each surface can be regulated independently in response to the environment \citep{Darwin_1898, Pospisilova_Solarova_1980, Smith_1981, Reich_1984, Mott_Oleary_1984}, amphistomatous leaves likely cope with these stresses by rapidly closing adaxial stomata when water supply cannot match evaporative demands. Instead, patterns in the British flora are at least consistent with the idea that adaxial stomata increase susceptibility to foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}. The cost of adaxial stomata may be greater in the shade because greater leaf wetness and lower ultraviolet light provide a more suitable microclimate for many foliar pathogens.

@@

These observations from the British flora strongly support the hypothesis that amphistomy can be used a proxy for open habitat in paleoenvironment reconstruction \citep{Carpenter_1994, Jordan_etal_2014, Carpenter_etal_2015}, but also point out previously unknown subtleties. These previous studies based their conclusions on data from Proteaceae, in which there is little quantitative variation in stomatal ratio; species are either completely hypostomatous ($\mathrm{SR_{propAd}} \approx 0$) or completely amphistomatous ($\mathrm{SR_{propAd}} \approx 0.5$). Stomatal ratio in British angiosperms is also bimodal \citep{Peat_Fitter_1994b}, but across many families there is also quantiative variation. Importantly, this means that quantitative variation in stomatal ratio may provide a more precise, quantiative indicator of vegetation type, rather than simply `open' or `closed'. A quantitative relationship between \el~and stomatal ratio has already been shown for herbaceous perennials \citep{Bucher_etal_2017}, but we now know that it holds among annuals (therophytes), subshrubs (chamaephytes), and, to a lesser extent, geophytes as well (Fig.~\ref{fig:SRmultReg}).

@@

Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio. The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSRa$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$). This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal. Regardless of the mechanism, this fact means that researchers may be able to use data from closely related species to improve paleoenvironment reconstruction.

@@

Variation in stomatal ratio is determined primarily by evolution of adaxial stomatal density and is coordinated with increases in total leaf stomatal density summed across both surfaces. Phylogenetic analyses show that changes in stomatal ratio and total stomatal density, especially in response to \el, are predominantly mediated by changes in adaxial stomatal density. This highly nonrandom pattern among British angiosperms mirrors evolutionary changes wrought by domesication \citep{Milla_etal_2013}; crops species tend to have higher adaxial stomatal density than their wild relatives. Note here that I am referring only to evolutionary variation in stomatal ratio among species; different processes may mediate within species variation or plastic responses.

There are at least two hypotheses that could explain why adaxial stomatal density is the most responsive. The first I refer to as the `real estate' hypothesis. In hypostomatous plants, the lower surface is already crowded with stomata, and hence plants must increase the real estate available for stomata by develping them on the upper surface whenever there is selection for greater stomatal density. When stomata are packed too densely on one surface, stomatal interference limits their functioning and hence may create a strong selective pressure for amphistomy \citep{Parlange_Waggoner_1970, Dow_etal_2014a}.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform shape stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ lifeform
  & \Sexpr{round(fitSRa$optpar, 2)}
  & \Sexpr{round(fitSRa$sigma2, 3)}
  & \Sexpr{round(cor(fitSRa$fitted.values, fitSRa$fitted.values + fitSRa$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRa)) + 2}
  & \Sexpr{round(fitSRa$logLik, 1)}
  & \Sexpr{round(fitSRa$aic, 1)}
  & \Sexpr{round(fitSRa$aic - min(aicSR), 1)} \\
    \el~+ lifeform
  & \Sexpr{round(fitSRb$optpar, 2)}
  & \Sexpr{round(fitSRb$sigma2, 3)}
  & \Sexpr{round(cor(fitSRb$fitted.values, fitSRb$fitted.values + fitSRb$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRb)) + 2}
  & \Sexpr{round(fitSRb$logLik, 1)}
  & \Sexpr{round(fitSRb$aic, 1)}
  & \Sexpr{round(fitSRb$aic - min(aicSR), 1)} \\
    \el
  & \Sexpr{round(fitSRc$optpar, 2)}
  & \Sexpr{round(fitSRc$sigma2, 3)}
  & \Sexpr{round(cor(fitSRc$fitted.values, fitSRc$fitted.values + fitSRc$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRc)) + 2}
  & \Sexpr{round(fitSRc$logLik, 1)}
  & \Sexpr{round(fitSRc$aic, 1)}
  & \Sexpr{round(fitSRc$aic - min(aicSR), 1)} \\
    lifeform
  & \Sexpr{round(fitSRd$optpar, 2)}
  & \Sexpr{round(fitSRd$sigma2, 3)}
  & \Sexpr{round(cor(fitSRd$fitted.values, fitSRd$fitted.values + fitSRd$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRd)) + 2}
  & \Sexpr{round(fitSRd$logLik, 1)}
  & \Sexpr{round(fitSRd$aic, 1)}
  & \Sexpr{round(fitSRd$aic - min(aicSR), 1)} \\
    1
  & \Sexpr{round(fitSRe$optpar, 2)}
  & \Sexpr{round(fitSRe$sigma2, 3)}
  & \Sexpr{0}
  & \Sexpr{length(coef(fitSRe)) + 2}
  & \Sexpr{round(fitSRe$logLik, 1)}
  & \Sexpr{round(fitSRe$aic, 1)}
  & \Sexpr{round(fitSRe$aic - min(aicSR), 1)} \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SRmultReg.pdf}}
\caption{The effect of light on stomatal ratio depends on Raunki\ae r life form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 1000 parametric bootstrap samples. Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visial clarity.} 
\label{fig:SRmultReg}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SD-light.pdf}}
\caption{Light-mediated evolution of stomatal ratio is mostly driven by increased adaxial (`upper') stomatal density (Panel A), whereas abaxial (`lower') stomatal density (Panel B) is similar across Ellenberg light indicator values (\el~$x$-axis). The violin plot shows stomatal density ($y$-axis, log-scale) as a function of \el. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the point indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Points outside the polygons are statistical outliers.} 
\label{fig:SD-light}
\end{figure}

>>>>

% Things to do before submission:
% Matt's notes
% Remove code to copy refs
% Github DOI?
% Finish abstract
% Finish citations (add citations only mentioned online to MS)
% Finish conclusions

@@

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}). Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SRmultReg}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distibuted across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. % Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

@@

Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio. The contributions of adaxial density, abaxial density, and their covariance are \Sexpr{round(varSDad / varSR, 2)}, \Sexpr{round(varSDab / varSR, 2)}, and \Sexpr{round(-2 * covSD / varSR, 2)}, respectively. This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.

Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}). Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ \Sexpr{pval2latex(exp(pvalSEM["logSDad ~ ellenberg_light"]))} and \Sexpr{pval2latex(exp(pvalSEM["logSDab ~ ellenberg_light"]))}, respectively). However, the regression of \el~on $\mathrm{sd_{ad}}$ was \Sexpr{round(fitSEM@ParTable$est[2] / fitSEM@ParTable$est[1], 1)}$\times$ that of \el~on $\mathrm{sd_{ab}}$ (\Sexpr{round(fitSEM@ParTable$est[2], 2)} versus \Sexpr{round(fitSEM@ParTable$est[1], 2)}). Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a \Sexpr{round(exp(fitSEM@ParTable$est[2]), 2)}-fold change in adaxial stomatal density versus a \Sexpr{round(exp(fitSEM@ParTable$est[1]), 2)}-fold change in abaxial stomatal density. The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $\Sexpr{pval2latex(exp(pvalSEM["logSDab ~~ logSDad"]))}). These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.

@@

The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance is not clear. Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work. Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}. However, none of these studies have accounted for the fact thats light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary hisory. Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods. As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Raunki\ae r life form; Fig.~\ref{fig:lf-light}). Nevertheless, both \el~and Raunki\ae r life form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms. These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.

@@

Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy. In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}). Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesies (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs). This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both annual and perennial forms). I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991}.

The prevalence of amphistomatous species in high light habitats supports the hypothesis that amphistomy is an adaptation to maximize photosynthetic rates by increasing CO$_2$ diffusion \citep{Jones_1985}. It is also evidence against the hypothesis that the principle fitness cost of amphistomy is water loss \citep{Darwin_1886, Foster_Smith_1986} or dehydration of pallisade mesophyll \citep{Buckley_etal_2015}, though these factors are likely very important in determining differential regulation of stomata on each surface. Since evaporative demand increases under high light, under these hypotheses we would expect plants in high light to be hypostomatous. Because stomatal conductances on each surface can be regulated independently in response to the environment \citep{Darwin_1898, Pospisilova_Solarova_1980, Smith_1981, Reich_1984, Mott_Oleary_1984}, amphistomatous leaves likely cope with these stresses by rapidly closing adaxial stomata when water supply cannot match evaporative demands \citep{Richardson_etal_2017}. Instead, patterns in the British flora are at least consistent with the idea that adaxial stomata increase susceptibility to foliar pathogens \citep{Gutschick_1984b, Mckown_etal_2014}. The cost of adaxial stomata may be greater in the shade because greater leaf wetness and lower ultraviolet light provide a more suitable microclimate for many foliar pathogens.

@@

These observations from the British flora partially support the hypothesis that amphistomy can be used a proxy for open habitat in paleoenvironment reconstruction \citep{Carpenter_1994, Jordan_etal_2014, Carpenter_etal_2015} but also point out previously unknown subtleties. These previous studies based their conclusions on data from Proteaceae, in which there is little quantitative variation in stomatal ratio; species are either completely hypostomatous ($\mathrm{SR_{propAd}} \approx 0$) or completely amphistomatous ($\mathrm{SR_{propAd}} \approx 0.5$). Stomatal ratio in British angiosperms is also bimodal \citep{Peat_Fitter_1994b}, but across many families there is also quantitative variation. Importantly, this means that quantitative variation in stomatal ratio may provide a more precise, quantiative indicator of vegetation type, rather than simply `open' or `closed'. A quantitative relationship between \el~and stomatal ratio has already been shown for herbaceous perennials \citep{Bucher_etal_2017}, but we now know that it holds among annuals (therophytes), subshrubs (chamaephytes), and, to a lesser extent, geophytes as well (Fig.~\ref{fig:SRmultReg}).

@@

Finally, phylogenetic information should improve inferences about paleoclimates because there is appreciable phylogenetic signal in stomatal ratio. The phylogenetic half-life of stomatal ratio evolution, after accounting for \el~and Raunki\ae r life form, is $\mathrm{log}(2) / \alpha = \Sexpr{round(log(2) / fitSRa$optpar, 1)}$ my (see Table~\ref{table:srAIC} for maximum likelihood estimates of $\alpha$, the return rate in the Ornstein-Uhlenbeck model of trait evolution). This lag time may indicate that evolving to the `optimum' is constrained by the shape of the fitness landscape \citep{Muir_2015} or that other unmeasured factors which affect stomatal ratio have some phylogenetic signal. Regardless of the mechanism, this fact means that researchers may be able to use data from closely related species to improve paleoenvironment reconstruction.

@@

Variation in stomatal ratio is determined primarily by evolution of adaxial stomatal density and is coordinated with increases in total leaf stomatal density summed across both surfaces. Note here that I am referring only to evolutionary variation in stomatal ratio among species; different processes may mediate within species variation or plastic responses. Phylogenetic analyses show that changes in stomatal ratio and total stomatal density, especially in response to \el, are predominantly mediated by changes in adaxial stomatal density. This highly nonrandom pattern among British angiosperms mirrors evolutionary changes wrought by domestication \citep{Milla_etal_2013}; crop species tend to have higher adaxial stomatal density than their wild relatives.

There are at least two hypotheses that could explain why adaxial stomatal density is the most responsive. The first I refer to as the `real estate' hypothesis. In hypostomatous plants, the lower surface is already crowded with stomata, and hence plants must increase the real estate available for stomata by developing them on the upper surface whenever there is selection for greater stomatal density. When stomata are packed too densely on one surface, stomatal interference limits their functioning and hence may create a strong selective pressure for amphistomy \citep{Parlange_Waggoner_1970, Dow_etal_2014a}.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ lifeform
  & \Sexpr{round(fitSRa$optpar, 2)}
  & \Sexpr{round(fitSRa$sigma2, 3)}
  & \Sexpr{round(cor(fitSRa$fitted.values, fitSRa$fitted.values + fitSRa$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRa)) + 2}
  & \Sexpr{round(fitSRa$logLik, 1)}
  & \Sexpr{round(fitSRa$aic, 1)}
  & \Sexpr{round(fitSRa$aic - min(aicSR), 1)} \\
    \el~+ lifeform
  & \Sexpr{round(fitSRb$optpar, 2)}
  & \Sexpr{round(fitSRb$sigma2, 3)}
  & \Sexpr{round(cor(fitSRb$fitted.values, fitSRb$fitted.values + fitSRb$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRb)) + 2}
  & \Sexpr{round(fitSRb$logLik, 1)}
  & \Sexpr{round(fitSRb$aic, 1)}
  & \Sexpr{round(fitSRb$aic - min(aicSR), 1)} \\
    \el
  & \Sexpr{round(fitSRc$optpar, 2)}
  & \Sexpr{round(fitSRc$sigma2, 3)}
  & \Sexpr{round(cor(fitSRc$fitted.values, fitSRc$fitted.values + fitSRc$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRc)) + 2}
  & \Sexpr{round(fitSRc$logLik, 1)}
  & \Sexpr{round(fitSRc$aic, 1)}
  & \Sexpr{round(fitSRc$aic - min(aicSR), 1)} \\
    lifeform
  & \Sexpr{round(fitSRd$optpar, 2)}
  & \Sexpr{round(fitSRd$sigma2, 3)}
  & \Sexpr{round(cor(fitSRd$fitted.values, fitSRd$fitted.values + fitSRd$residuals) ^ 2, 2)}
  & \Sexpr{length(coef(fitSRd)) + 2}
  & \Sexpr{round(fitSRd$logLik, 1)}
  & \Sexpr{round(fitSRd$aic, 1)}
  & \Sexpr{round(fitSRd$aic - min(aicSR), 1)} \\
    null
  & \Sexpr{round(fitSRe$optpar, 2)}
  & \Sexpr{round(fitSRe$sigma2, 3)}
  & \Sexpr{0}
  & \Sexpr{length(coef(fitSRe)) + 2}
  & \Sexpr{round(fitSRe$logLik, 1)}
  & \Sexpr{round(fitSRe$aic, 1)}
  & \Sexpr{round(fitSRe$aic - min(aicSR), 1)} \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SRmultReg.pdf}}
\caption{The effect of light on stomatal ratio depends on Raunki\ae r life form. Greater Ellenberg light indicator values (\el) are associated with greater stomatal ratio ($\mathrm{SR_{even}}$) in therophytes, hemicryptophytes, and chamaephytes but not geophytes and phanerophytes. The maximum likelihood slope from phylogenetic regression is given with statistical significance based on 1000 parametric bootstrap samples. Numbers in parentheses next to Raunki\ae r life form are the sample sizes in the final dataset. Estimated slopes (solid line) and 95\% bootstrapped confidence intervals (gray polygon between dashed lines) are plotted against raw data. Points have been jittered for visual clarity.} 
\label{fig:SRmultReg}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics[width=0.5\textwidth]{figures/figure_SD-light.pdf}}
\caption{Light-mediated evolution of stomatal ratio is mostly driven by increased adaxial (`upper') stomatal density (Panel A), whereas abaxial (`lower') stomatal density (Panel B) is similar across Ellenberg light indicator values (\el~$x$-axis). The violin plot shows stomatal density ($y$-axis, log-scale) as a function of \el. The width of the grey polygons indicates the density of data. Length of grey polygon indicate the range of the data; the dot indicates the median; the thick lines indicate the 0.25 and 0.75 quantiles. Points outside the polygons are statistical outliers.} 
\label{fig:SD-light}
\end{figure}

### 2017-07-13 ms/ms.tex
edited ms

% NOTES

%  papers on Ellenberg light value:
%  Bartelheimer and Poschlod 2015
%  Diekmann, 2003

@@

I tested whether Raunki\ae r life form was associated \el~values among British angiosperms using ANOVA with Type-2 sum of squares. I did not use phylogenetic ANOVA for this test because there was no phylogenetic signal in the regression fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}. See the R code accompanying this paper for further detail. I predicted that species with faster life histories, especially therophytes (annuals), would have greater \el~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.

@@

Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$). Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \el~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}). Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals did not overlap zero (Fig.~\ref{fig:SRmultReg}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distibuted across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

>>>>

@@

I tested whether Raunki\ae r life form was associated \el~among British angiosperms using ANOVA with Type-2 sum of squares. I did not use phylogenetic ANOVA for this test because there was no phylogenetic signal in the regression fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}. See the R code accompanying this paper for further detail. I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes, which are mostly long-lived trees.

@@

Ellenberg light indicator values (\el) differed significantly among life forms (Fig.~\ref{fig:lf-light};ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$). Therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (large woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}).

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}). Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals did not overlap zero (Fig.~\ref{fig:SRmultReg}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distibuted across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).
